A spatio-temporal analysis of the environmental correlates of COVID-19 incidence in Spain

### 2020-04-20 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain-Title-Page.tex
Revised abstract before submission

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={A spatio-temporal analysis of the environmental correlates of COVID-19 incidence in Spain},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

% Pandoc header
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

  \title{A spatio-temporal analysis of the environmental correlates of COVID-19
incidence in Spain}
    \author[Some School]{Author A\corref{1}}
   \ead{author.a@example.com} 
    \author[Some Institute]{Author B}
   \ead{author.b@example.com} 
      \address[Some School]{School of the Things}
    \address[Some Institute]{Institute of Everything}
    \address[Departamento de Economia]{}
      \cortext[1]{Corresponding Author}

  \begin{abstract}
  Spreading with astonishing speed, the novel SARS-CoV2 has swept the
  globe, causing enormous stress to health systems and prompting social
  distance guidelines and mandates to arrest its progress. While there is
  encouraging evidence that early public health interventions have slowed
  the spread of the virus, this has come at a high cost as the global
  economy is brought to its knees. How and when to ease restrictions to
  movement hinges in part on the question whether SARS-CoV2 will display
  seasonality associated with variations in temperature, humidity, and
  hours of sunshine. In this research, we address this question by means
  of a spatial analysis of the incidence of COVID-19 in the provinces in
  Spain. Use of a spatial Seemingly Unrelated Regressions (SUR) approach
  allows us to model the incidence of reported cases of the disease per
  100,000 population, as a function of temperature and humidity, while
  controlling for GDP per capita, population density, percentage of older
  adults in the population, and presence of mass transit systems. An
  interesting aspect of the spatial SUR approach is that it models
  incidence as a contagion process. Our results indicate that incidence of
  the disease is lower at higher temperatures and higher levels of
  humidity, although coefficients for this variable are significant only
  in some equations. Sunshine, in contrast, displays a positive
  association with incidence of the disease. Our control variables also
  yield interesting insights. Higher incidence is associated with higher
  GDP per capita and presence of mass transit systems in the province; in
  contrast, population density and percentage of older adults display
  negative associations with incidence of COVID-19.
  \end{abstract}

@@

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

From a small outbreak linked to a live animal market at the end of 2019
to a global pandemic in a matter of weeks, the SARS-CoV2 virus has
threatened to overrun health systems the world over. In efforts to
contain the spread, numerous governments in many nations and regions
have either recommended or mandated social distancing measures, and as
of this writing, millions of people in five continents shelter in place.
There are encouraging signs that these measures have arrested the spread
of the virus where they have been implemented, and have thus helped to
keep a bad situation from becoming even worse (e.g., 2020). However,
this has come at a high cost, and the consequences for all spheres of
the economy, social, and cultural life have been dire (e.g., Fernandes,
2020; Luo and Tsang, 2020). As a result, there is a sense of urgency to
anticipate the progression of the pandemic, in order to plan for
progressive lifting of restrictions to movement and social contact
(e.g., Kissler et al., 2020). Needless to say, this has become a
delicate, and politically charged, balancing act between public health
and the economy (Gong et al., 2020).

A salient question in the debate on how and when to ease social
distancing measures is whether the virus will display seasonal
variations. Earlier, diverse studies have shown the effect of
temperature and humidity on the incidence of influenza (e.g., Mäkainen
et al., 2009; Jaakkola et al., 2014; Kudo et al., 2019). Jaakkola et
al.~(2014), for example, found that a decrease of temperature and
absolute humidity precedes the onset of symptoms of influenza A and B
viruses by 3 days in places where the temperature is low. After the
2002-2004 outbreak of SARS, researchers also began to investigate the
relationship between these factors and SARS-CoV. In this way, Casanova
et al.~(2010) used two surrogates, namely the gastroenteritis (TGEV) and
mouse hepatitis viruses (MHV), to find that virus inactivation was more
rapid at temperatures of 20C than 4C, and at 40C than 20C; in terms of
humidity, these researchers reported that survival of the virus was
lower at moderate relative humidity levels. In a similar vein, but
working directly with SARS-CoV, Chan et al.~(2011) found that viability
of the virus was lost at temperatures higher than 38C and relative
humidity superior to 95\%.

While existing research on similar pathogens suggests that SARS-CoV is
more stable and potentially easier to transmit in conditions of low
temperature and low humidity, it is far from certain that this will also
be the case with the novel SARS-CoV2. If such is the case, there is the
possibility of easing restrictions to social contact as the weather
warms; however, weeks or possibly months of costly measures could become
undone if not, and the restrictions are lifted prematurely. Not
surprisingly, given the stakes involved, this issue has already
triggered a lively debate.

Some of what is thought about the possible seasonality of COVID-19 is
based on analogies to the patterns of other known respiratory viruses.
However, de Ángel Solá et al.~(2020) note that ``not all seasonal
respiratory viruses experience the same spatiotemporal patterns''
(section 4). This urges caution when extrapolating from known viruses,
and indeed, the evidence in this respect is inconclusive. At a global
scale, de Ángel Solá et al.~(2020) see less risk in the Caribean Basin;
however, Coelho et al.~(Coelho et al., 2020) warn that at least during
the exponential phase, expansion of the virus is not driven by climate.
Similarly, whereas Araujo and Naimi (2020) argue that spread of
SARS-CoV2 will likely be constrained by climate, Harbert et al.~(2020)
remain unconvinced that spatial modelling can currently discriminate the
distribution of the disease on the basis of climate, at least in the
United States. Yao et al.~(2020), examined data from China and came to
the conclusion that neither temperature nor ultraviolet indices had an
association with transmission of COVID-19. This is despite previous
research that has linked less exposure to UVB radiation to higher
prevalence and severity of acute respiratory tract infections
(Zittermann et al.~2016; Dąbrowska-Leonik et al.~2018; Dinlen et
al.~2016; Mathyssen et al.~2017; Esposito and Lelii 2015; Jat 2017;
Moriyama, Hugentobler, and Iwasaki 2020). To further complicate matters,
much of the relevant work has yet to be peer-reviewed and therefore is
open to change (see for example the challenge of Harbert et al. (2020)
to Araujo and Naimi (2020)). In the United States, the National Academy
of Sciences, Engineering, and Medicine was engaged by the Office of the
Executive for guidance on this matter (see National Academies of
Sciences, Engineering and Medicine, 2020). Their conclusion summarizes
the situation well (see p.~6): ``Some limited data support a potential
waning of cases in warmer and more humid seasons, yet none are without
major limitations\ldots Additional studies as the SARS-CoV-2 pandemic
unfolds could shed more light on the effects of climate on
transmission.''

With the above considerations in mind, our objective with this paper is
to contribute to the knowledge basis regarding the spread of COVID-19
and the influence of environmental factors, particularly temperature,
humidity, and sunshine. We adopt a population health approach, and
report results from a spatial model of the incidence of COVID-19 in
fifty provinces in Spain, one of the countries hardest hit by the
pandemic. We combine data on reported cases of the disease with
metereological information, to create a spatio-temporal dataset covering
a period of 30 days. We then join this dataset with provincial-level
economic and demographic information to act as controls to our key
environmental variables. These data are analyzed using a spatial
Seemingly Unrelated Regressions (SUR) approach, which allows us to model
incidence of COVID-19 as a spatial contagion process.

The results provide evidence of the effect of temperature, humidity, and
sunshine on the incidence of COVID-19. The clearest result with respect
to these variables is a lower incidence of COVID-19 at higher
temperatures and levels of humidity, while the opposite happens with
respect to hours of sunshine. Our control variables also provide some
intriguing insights. Higher incidence is associated with higher GDP per
capita and presence of mass transit systems in the province; in
contrast, population density and percentage of older adults display
negative associations with incidence of COVID-19. The results of this
analysis provide support to the hypothesis of seasonality of the novel
SARS-CoV2, and should be of interest to public health officials and
policy makers grappling with the question of the trajectory of the
pandemic.

Please note that this paper is prepared as a reproducible research
document. The source R Markdown document, as well as all data and code
needed to reproduce/review/extend the analysis can be obtained from the
following anonymous Drive folder:

https://drive.google.com/open?id=1d\_40N\_QXo2Fl14r3T3CHs84IED-B1rV\_

\hypertarget{context-and-data}{%
\section{Context and Data}\label{context-and-data}}

\hypertarget{covid-19-in-spain}{%
\subsection{Covid-19 in Spain}\label{covid-19-in-spain}}

The first reported case of COVID-19 in Spain was on January 31th, when a
German tourist in the Canary Islands tested positive for the virus.
However, it was still a few weeks before the first domestic case was
reported, on February 27th in Sevilla province (Andalusia). In a short
period of time, after this relatively slow start, the outbreak flared.
By March 11th the World Health Organization (WHO) declared COVID-19
officially a pandemic. This declaration marked a turning point for the
public in Spain too. As of March 13th, the number of cases of COVID-19
reported in Spain was 4,473, with a majority of cases (1,990)
concentrated in Madrid: these numbers were at the time the worst
outbreak in Europe, after Italy. In response to the situation, on March
13th the Spanish National Government declared a state of emergency, to
go into effect on Saturday March 14th. As part of the state of emergency
restrictions to most activities were imposed, with the exception of
essential services (e.g.~food, health) and some economic subsectors of
industry and construction. A few days later, on March 17th, Spain closed
its lands borders to allow entry only to returnee nationals and
permanent residents. The lockdown was further hardened between March
30th and April 12th (including the Easter weekend of April 10th-12th)
and during this period only essential activities were allowed. During
this period, there was a dramatic reduction in overall mobility, both
within provinces as between
\footnote{https://www.mitma.gob.es/ministerio/covid-19/evolucion-movilidad-big-data/movilidad-provincial}.

\hypertarget{variables}{%
\subsection{Selection of Variables}\label{variables}}

The global emergence of infectious diseases is mostly driven by
environmental, ecological, and socio-economic factors (Jones et al.,
2008). In the case of SARS-CoV2, the ecological factors include the
interaction between humans and wildlife. Once transmission of a disease
begins to happen between humans, socio-economic and environmental
factors become increasingly important. As noted in the introduction, the
focus of the paper is on environmental variables, namely temperature,
humidity, and sunshine. These variables have been implicated in the
viability and ease of transmission of similar viruses. In addition to
these variables, we also aim to use a set of controls, in the form of
specific socio-economic and demographic characteristics of each
province.

The first variable that we consider is GDP per capita. Much has been
said about globalization and the spread of infectious
disease\footnote{As the Globe and Mail, Canada's paper of record, put it in relation to the SARS outbreak in 2003: "Globalization means that if someone in China sneezes, someone in Toronto may one day catch a cold" (Editorial, March 28, 2003, p. A18)}.
The growth in global connections has presented a challenge to spatial
approaches in the initial stages of disease management, when the cause
of a disease may still be unclear, but the plane has already departed
(Zhou and Coleman, 2016). In reference to the earlier outbreak of SARS,
van Wagner (Van Wagner, 2008) chronicles how Toronto's status as a
global city turned out to be a vulnerability in this respect. In our
case, we think of GDP per capita as a marker of a region's relative
position in a network of global cities, and its potential to be further
ahead in the trajectory of the pandemic. Furthermore, wealthier regions
also tend to concentrate more activities that produce non-traded goods,
including building and construction (Hallet, 2002). Therefore, it is
possible that wealthier regions remain relatively more active even
during a lockdown. On the other hand, it is also possible that less
wealthy regions have a higher proportion of workers in manual
occupations that cannot telework, and therefore have more difficulties
complying with shelter-in-place orders.

The percentaje of older adults (over 65) is the second variable that we
consider as a control. Early evidence regarding COVID-19 suggests that
the case rate mortality is higher at older ages (e.g.~The Novel
Coronavirus Pneumonia Emergency Response Epidemiology Team, 2020).
However, it is not clear that a relatively large population of older
adults necessarily translates into higher transmission of the infection.
The tool of choice in containing the spread of the disease has been
social distancing. In this respect, the evidence from the field of
transportation is that older adults tend to travel less frequently, for
shorter distances, and have higher rates of immobility (e.g., Roorda et
al., 2010; Morency et al., 2011; Sikder and Pinjari, 2012). In other
words, many older adults are, whether by preference or otherwise,
already in a form of social isolation. Social distancing during the
pandemic may actually reinforce that condition for them, as suggested by
the analysis of age-structured social contact in India, China, and Italy
of Sing and Adhikari (2020). Accordingly, our expectation is that
provinces with higher percentages of older adults will tend to have
similarly low levels of social contact, particularly since the
age-structured matrix of social contact in Spain is similar to Italy
(Prem et al., 2017).

The third population variable the we consider is population density,
since it directly affects the contact patterns and contact rates between
individuals in a population (Hu et al., 2013). The evidence available
suggests a positive relationship between the transmission of COVID-19
and population density (e.g.~cumulative incidence in urban areas like
NYC). For this reason, we anticipate a positive relationship between
population density and the incidence of the disease.

The last control variable is the presence of mass transit systems in a
province. Every province in Spain offers some form of public
transportation, however only five provinces have higher order systems of
mass mobility (e.g.~metro or subway), namely Barcelona, Madrid, Sevilla,
Valencia, and Bizkaia. Public transportation has been hypothesized to
relate to the spread of contagious disease by some researchers using
agent-based approaches and simulation (e.g., Perez and Dragicevic, 2009;
Wang et al., 2010), and while we find scant evidence of a link in the
literature, the idea is intuitively appealing. After all, unlike the
isolation that a car offers to travellers, most mass transit system are
cauldrons of social contact.

\hypertarget{data}{%
\subsection{Data}\label{data}}

Our dataset includes information about the daily number of cases of
COVID-19 reported in Spain at the provincial level (NUTIII in Eurostat
terminology) for the period between March 13th and April 11th,
inclusive. For our purposes, we consider positive cases reported, but
excluding symptomatic cases diagnosed by a doctor without a Polymerase
Chain Reaction (PCR) test. The Spanish National Government publishes
periodic updates at the regional level (NUTII) and the information is
also released at the provincial level as part of a collaborative project
by geovoluntarios.com\footnote{https://www.geovoluntarios.org/},
ProvidencialData19\footnote{https://www.datoscovid.es/pages/providencialdata19},
and ESRI España. This information is compiled from several sources,
mainly the official web pages of the Spanish regional goverments, as
documented in Centro de Datos
Covid-19\footnote{https://www.datoscovid.es/pages/sobre-la-iniciativa}.
In addition, we consider two sets of explanatory variables. The first
one, and the focus of this research, is set of two environmental
variables, namely temperature and humidity, which are collected from
official sources (i.g., \emph{AEMET}, the state meteorology agency, and
\emph{MAPA}, the ministry of agriculture, fisheries, and food). The
second set provides some relevant controls for multivariate analysis,
and refers to economic and demographic attributes of the province (also
collected from official sources, i.e., INE, the national statistics
institute). Table \ref{tab:descriptive-statistics} shows the descriptive
statistics and the provenance of the data.

The spatial and temporal coverage of the data is as follows. Our dataset
begins on March 13, which is the first date when every province had
reported at least one case of COVID-19, and continues until April 11,
for a period of 30 days. The unit of analysis is the province. Provinces
are the equivalent of states, and are embedded in Autonomous
Communities. As an example, Cataluña is an Autonomous Community and
consists of four provinces, namely Barcelona, Gerona, Lerida, and
Tarragona. The size of the provinces is relatively large, as seen in
Table \ref{tab:descriptive-statistics}. The smallest province is
\(1,978.12km^2\) (this is smaller than Rhode Island in the US) and the
largest province is \(21,767.93km^2\) (slightly smaller than New Jersey
in the US). While this is a relatively large degree of spatial
aggregation, reporting on COVID-19 is inconsistent at smaller
geographies, or cases are not reported at that level at all. The
analysis must therefore be considered ecological.

\begin{table}

\caption{\label{tab:descriptive-statistics}\label{tab:descriptive-statistics} Descriptive statistics}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{l>{\raggedright\arraybackslash}p{15em}rrrr>{\raggedright\arraybackslash}p{5em}}
\toprule
Variable & Note & Min & Mean & Max & SD & Source\\
\midrule
\rowcolor{gray!6}  COVID-19 Incidence & Incidence in reported cases of SARS-19 per 100,000 people & 0.38 & 153.61 & 1149.36 & 186.23 & ProvidencialData19\\
Area & Area of province in sq.km & 1978.12 & 10118.79 & 21767.93 & 4.77 & INE\\
\rowcolor{gray!6}  GDPpc & GDP per capita in €1,000s & 16.67 & 22.51 & 36.00 & 4805.98 & INE\\
Older & Percentage of people aged 65 and older in the province & 15.16 & 21.03 & 31.36 & 3.95 & INE\\
\rowcolor{gray!6}  Population Density & Population density in the province in people per sq.km & 8.60 & 140.04 & 829.76 & 181.25 & INE\\
\addlinespace
Mean Temperature & Mean temperature in province by date, in C & 1.00 & 12.18 & 23.20 & 3.67 & AEMET\\
\rowcolor{gray!6}  Humidity & Relative humidity in province by date & 2.00 & 77.82 & 100.00 & 10.37 & MAPA\\
\bottomrule
\multicolumn{7}{l}{\textit{Note: }}\\
\multicolumn{7}{l}{ProvidencialData19: https://www.datoscovid.es/pages/providencialdata19}\\
\multicolumn{7}{l}{INE (Instituto Nacional de Estadistica): https://www.ine.es/}\\
\multicolumn{7}{l}{AEMET (Agencia Estatal de Meteorologia): http://eportal.mapa.gob.es}\\
\multicolumn{7}{l}{MAPA (Ministerio de Agricultura, Pesca y Alimentacion): http://eportal.mapa.gob.es}\\
\end{tabular}}
\end{table}

An important aspect of working with environmental data such as
temperature and humidity is the incubation period of the disease. Lauer
et al.~(2020) report for the case of COVID-19 a median incubation period
of 5.7 days (with a confidence interval between 4.9 to 7.8 days). The
vast majority of cases (95\%) develop symptoms between 2.6 days (CI, 2.1
to 3.7 days) and 12.5 days (CI, 8.2 to 17.7 days). For this reason, we
judge it best to use lagged values of the environmental variables. We
test different time lags as follows. We consider a lagged 8-day average,
from date-minus-12 to date-minus-5 days (hereafter \emph{lag8}).
Secondly, we consider a lagged 11-day average, from date-minus-12 to
date-minus-2 days (hereafter \emph{lag11}). Finally, to account for the
likely duration of incubation, we consider a lagged 11-day
\emph{weighted} average, from date-minus-12 to date-minus-2 days
(hereafter \emph{lag11w}). The weights for this variable are calculated
using the parameters of the log-normal distribution reported by Lauer et
al.~(2020), i.e., a log-mean of 1.621 and a log-standard deviation of
0.418. With these weights, the environmental variables at date-minus-2
and date-minus-12 days are weighted as 0.041 and 0.009 respectively,
whereas the environmental variables at date-minus-5 days are weighted as
0.194.

\hypertarget{methods}{%
\section{Methods: the Spatial SUR Model}\label{methods}}

The Seemingly Unrelated Regression equations model (SUR hereafter) is a
multivariate econometric model used in different fields when the
structure of the data consisits of cross-sections for different time
periods. The basis of this approach is well-known since the initial
works of Zellner (1962), and has become a popular methodology included
in several econometrics textbook (e.g., Greene, 2003). To our knowledge,
Anselin (1988) was the first author to discuss SUR from a spatial
perspective. In his landmark text, Anselin discussed a model made of
``an equation for each time period, which is estimated for a cross
section of spatial units'' (p.~141). From this milestone, a large body
of research has developed to extend the classical SUR into a spatial
framework (e.g., Rey and Montouri, 1999; Lauridsen et al., 2010; Le
Gallo and Dall'Erba, 2006; López et al., 2017).

The classical SUR model without spatial effects (from here, SUR-SIM) is
a stack of equations as follows:

\begin{equation}
\label{eq:sur-sim}
\begin{bmatrix}
y_1 \\ y_2 \\ \vdots \\ y_T
\end{bmatrix}
=
  \begin{bmatrix}
X_1 & 0 & \cdots & 0 \\ 0 & X_2 & \cdots & 0 \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \cdots & X_T
\end{bmatrix}
\
\begin{bmatrix}
\beta_1 \\ \beta_1 \\ \vdots \\ \beta_T
\end{bmatrix}
+
\begin{bmatrix}
\epsilon_1 \\ \epsilon_2 \\ \vdots \\ \epsilon_T
\end{bmatrix}
\end{equation}

\noindent where \(y_{t}=(y_{1t},...,y_{Nt})\) is a \(N \times 1\)
vector, and in our case \(y_{st}\) is the incidence ratio in the
province \(s\) (\(s=1,...,N\)) the day \(t\) \((t=1,...,T)\); \(X_t\) is
a \(N \times k_t\) matrix of the \(k_t\) independent variables, with
associated vector of coefficients \(\beta_t\),;
\(\beta_t=(\beta_{1t},...,\beta_{Nt})\) is a vector of coefficients and
\(\epsilon_t=(\epsilon_{1t},...,\epsilon_{Nt})\) is the vector of
residuals.

A key feature of the SUR model is the dependence structure among the
vectors of residuals, namely:

\begin{equation}
\label{eq:sur-err}
E[\epsilon_t \epsilon'_{t'}]=\sigma_{tt'}
\end{equation}

Note that this specification is very flexible, in that it allows changes
in the coefficients \(\beta_{it}\) in order to modulate the effect of
\(X^i_{.t}\) on \(y_t\). This flexibility can be reduced and it is
posible to impose restrinctions considering some \(\beta\) coefficients
as being constant over time. In this case, we can reformulate the
coefficients expression
\(\beta_t = (\beta_{1}, \cdots, \beta_{r-1}, \beta_{r}, \beta_{r+1}, \cdots, \beta_{Nt})\)
to restrict the first \(r\) coefficients to be constant across
equations. This is equivalent to specifying some effects to be invariant
over time.

Equation (\ref{eq:sur-sim}) can be rewriten in compact form:

\begin{equation}
\bf{Y} = \bf{X} \beta + \epsilon
\label{eq:sur-sim-block}
\end{equation}

\noindent where \(\bf{Y}\) is now a vector of dimension \(NT \times 1\),
\(\bf{X}\) is a block-diagonal matrix \(NT \times K\) (with
\(K = \sum_t{k_t}\)) and \(\epsilon\) is an \(NT \times 1\) vector.
Using the Kronecker product notation (\(\otimes\)), the error matrix
structure is expressed concisely as:

\begin{equation}
E[\epsilon \epsilon']=\Sigma \otimes I_N ; \ \Sigma=(\sigma_{tt'})
\end{equation}

As is the case with cross-sectional data, it is possible to test the
residuals of Model (\ref{eq:sur-sim-block}) for spatial autocorrelation,
and several tests have been developed to test the null hypothesis of
spatial independence (see López et al., 2014). When the null hypothesis
is rejected, several alternative specifications have been proposed to
include spatial effects (Anselin, 1988, see also 2016). In this paper we
consider a spatial SUR model that incorporates a spatial lag of the
dependent variable as an explanatory factor. Spatial analytical
approaches were used to understand contagion-difussion processes in the
case of infectious disease in general (e.g., Cliff et al., 1998) and the
2003-2004 SARS outbreak in particular (e.g., Meng et al., 2005; Cao et
al., 2010). While we are mindful of the same caveat that the novel
SARS-CoV2 may not follow the patterns of previous diseases, there is
still evidence from the United States that COVID-19 displays spatial
patterns that are consistent with a diffusion process (Desjardins et
al., 2020). For this reason, the spatial lag model is appropriate to
model incidence of COVID-19 geographically, since it accounts for
potential spatial patterns that result from a process of contagion, as
explained next.

The stack expresion for the SUR model with a spatially lagged dependent
variable (SUR-SLM) is as follows:

\begin{equation}
\label{eq:sur-slm}
\begin{aligned}
\bf{AY} = \bf{X} \beta + \epsilon \\
\epsilon =N(0,\bf{\Sigma})
\end{aligned}
\end{equation}

\noindent where \(\bf{A} =I_{TN}-\bf{\Gamma} \otimes W\) is the
spatially lagged dependent variable, and
\(\bf{\Gamma} = diag(\rho_1, \cdots, \rho_T)\).

This specification assumes that outcome (\(y_{st}\)) at location \(s\)
and time \(t\) is partially determined by the weighted average
(\(Wy_{st}\)) of the outcome in neighboring provinces, with neighborhood
defined by matrix \(W\) of spatial weights. In other words, the
spatially lagged dependent variable represents a process of contagion,
where the disease in neighboring provinces can spillover in a spatial
way. The coefficients of the spatially lagged variable are estimated for
each time period \(\rho_t\) and identify the intensity and the sign of
the contagion effect. It is possible test the null hypothesis of
identical levels of spatial dependence (\(\rho_i=\rho_j, \forall i,j\)).
The correspond Wald test is available in the \texttt{R} package
\texttt{spsur}.

The SUR-SLM model can be estimated using maximum likelihood (López et
al. (2014)) or instrumental variables (Mínguez et al. (2019)).

A note regarding the interpretation of the model is in order. It is
well-known that coefficients in linear regression models are partial
derivatives of the dependent variable with respect to the independent
variables, and therefore directly give the marginal effects or rates of
change. Spatially lagged models, however, are no longer linear. The
intuition behind the non-linearity is that the spatial lag expands the
information set to include information from neighbouring regions: in
other words, the value of an explanatory variable in a spatial unit can
have influence in other spatial units via the spatial lag. This makes
interpretation of the coefficients less straightforward but also richer
(Golgher and Voss, 2016). The results of LeSage (2009) for
cross-sectional spatial lag models can be extended to the spatial SUR
framework. Note that, according to Elhorst (2014), the partial
derivatives have the following interpretation: if an explanatory
variable (\(X_k\)) in a particular province changes, not only the
incidence rate in that province changes, also incidence rates in other
provinces change via the contagion effect. Therefore, a change in
\(X_k\) in a particular province has a \emph{direct effect} on that
province, but also an \emph{indirect effect} on neighbouring provinces.
In this way, the \(i\)th diagonal element of the matrix of partial
derivatives represents the direct effect on the \(i\)th province,
whereas the non-diagonal elements of \(i\)th row of the matrix of
partial derivatives represent the indirect effects on other provinces.
In order to obtain a global indicator, the direct effect is calculated
as the mean of the diagonal elements and captures the average change in
incidence ratio caused by the change in \(X_k\). Likewise, a global
indicator of the indirect effects is calculated as the mean of the
non-diagonal elements. The total effect is the sum of direct and
indirect effects. Finally, note that if \(\rho_k = 0\), the indirect
effects are 0 and the direct effects are equal to \(\beta_kt\).

\hypertarget{analysis}{%
\section{Analysis}\label{analysis}}

\hypertarget{eda}{%
\subsection{Exploratory Data Analysis}\label{eda}}

Figure \ref{fig:weekly-average-incidence-map} shows the geographical
variation in the incidence of COVID-19 in Spain, as well as the temporal
progression of the disease in weekly averages. Our analysis begins on
March 13, which is the first date when every province had reported at
least one case of COVID-19. It can be seen that the highest incidence at
this early date was in the provice of Álava, in the North of Spain.
While not the most populous province, with a population of only 331,549,
Álava has the highest GDP per capita of all provinces. Vitoria, its main
city, is the capital of the Basque Country and has been the focus of
efforts, along with San Sebastian and Bilbao, to develop a ``Global
Basque City'' (Meijers et al., 2008). The other early focus of the
pandemic in Spain was Madrid, which is the most populous province in the
country and has the second highest GDP per capita after Álava. The early
outbreaks in these two provinces can be traced throughout the
progression of the pandemic over time, although by the end of the period
under consideration, other provinces had matched and even surpased their
incidence rates, including Segovia and Soria to the north of Madrid, and
Ciudad Real and Albacete to the south.

\begin{figure}
\centering
\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/weekly-average-incidence-map-1.pdf}
\caption{\label{fig:weekly-average-incidence-map}Mean weekly incidence
of COVID-19 by province, in reported cases by 100,000 people}
\end{figure}

Figure \ref{fig:descriptives-temperature} shows the distribution of the
environmental variables in Spain. For ease of visualization we aggregate
the provinces by Autonomous Community. Each box-and-whisker in the
figure represents the distribution of the variable for a community over
the 30-day period. In the plot, the communities have been sorted by
latitude, so that Principado de Asturias is the northernmost community,
and Canarias the southernmost. As seen in the figure, there is a
relatively wide range of values both within and between provinces over
the 30-day period examined. The top panel of the figure shows the
distribution of mean temperatures. The lowest mean temperature for a
community on any given day was 2.8C, and the highest 22.4C, for a range
of approximately 20 degrees. Likewise, there is a great deal of
variability in humidity, as seen in the middle panel of the figure,
where the lowest mean humidity for any community is 48.3 and the highest
is 99.6. Finally, the bottom panel displays mean daily hours of sunshine
in the community. This variable displays the most variability within
communities over time, but remains relatively constant across
communities. It is important to note that these are summaries by
community, and the actual values of these variables for the provinces
display somewhat more variability.

\begin{figure}
\centering
\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/descriptives-temperature-1.pdf}
\caption{\label{fig:descriptives-temperature} Distribution of mean
temperatures and humidities in the Autonomous Communities in Spain
between March 12, 2020 and April 11, 2020. The Autonomous Communities
have been sorted by latitude, with communities to the left being the
northermost, and to the right the southernmost}
\end{figure}

>>>>

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={A spatio-temporal analysis of the environmental correlates of COVID-19 incidence in Spain},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

% Pandoc header

@@

  \title{A spatio-temporal analysis of the environmental correlates of COVID-19
incidence in Spain}
    \author[McMaster University]{Antonio Paez\corref{1}}
   \ead{paezha@mcmaster.ca} 
    \author[Universidad Politecnica de Cartagena]{Fernando A. Lopez}
   \ead{fernando.lopez@upct.es} 
    \author[Departamento de Economia]{Tatiane Menezes}
   \ead{tatiane.menezes@ufpe.br} 
    \author[Nucleo de Pesquisa]{Renata Cavalcanti}
   \ead{renata.vcsantos@gmail.com} 
    \author[Nucleo de Pesquisa]{Maira Galdino da Rocha Pitta}
   \ead{mgrpitta@ufpe.br} 
      \address[McMaster University]{School of Geography and Earth Sciences, McMaster University, 1281 Main
St W, Hamilton, ON, L8S 4K1, Canada}
    \address[Universidad Politecnica de Cartagena]{Departamento de Metodos Cuantitativos, Ciencias Juridicas, y Lenguas
Modernas, Universidad Politecnica de Cartagena, Calle Real Numero 3,
30201, Cartagena, Murcia, Spain}
    \address[Departamento de Economia]{Departamento de Economia, Universidade Federal de Pernambuco, Av dos
Economistas, s/n - Cidade Universitária, Recife - PE, 50670-901, Brasil}
    \address[Nucleo de Pesquisa]{Núcleo de Pesquisa em Inovação Terapêutica NUPIT / UFPE, Av.
Prof.~Moraes Rego, 1235 - Cidade Universitária, Recife, PE, CEP
50670-901, Brazil}
      \cortext[1]{Corresponding Author}

  \begin{abstract}
  Early in 2020 the novel SARS-CoV2 swept the globe, disrupting health
  systems and the economy. Public health interventions have slowed the
  spread of the virus, albeit at a high cost. How and when to ease
  restrictions to movement hinges in part on whether SARS-CoV2 will
  display seasonality due to variations in temperature, humidity, and
  hours of sunshine. Here, we address this by means of a spatio-temporal
  analysis of the incidence of COVID-19 in Spain. Use of spatial Seemingly
  Unrelated Regressions (SUR) allows us to model the incidence of reported
  cases of the disease per 100,000 population as an interregional
  contagion process, in addition to a function of temperature, humidity,
  and sunshine. In the analysis we also control for GDP per capita,
  percentage of older adults in the population, population density, and
  presence of mass transit systems. The results support the hypothesis
  that incidence of the disease is lower at higher temperatures and higher
  levels of humidity. Sunshine, in contrast, displays a positive
  association with incidence of the disease. Our control variables also
  yield interesting insights. Higher incidence is associated with higher
  GDP per capita and presence of mass transit systems in the province; in
  contrast, population density and percentage of older adults display
  negative associations with incidence of COVID-19.
  \end{abstract}

@@
